White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-12-05 papers/wdmerger_II/paper.tex
Update the burning stability comments

It is true that when our resolution is low enough, there is a floor on the size
of a hotspot, possibly making such a detonation more likely. However, this is an
unavoidable consequence of the low resolution: it is the correct result of the
simulation that was performed. If the results do not match what occurs at higher
resolution, then the simulation is not converged and the results are not reliable.
However, it may also be the case that a higher resolution simulation will yield
similar results, for example because even at the higher resolution, the physical
size of the hotspot stays the same. One cannot appeal to the numerical instability
criterion alone to make a decision in this context.

Furthermore, we find that \textit{all} of the detonations we have seen at low
and moderate resolution inevitably violate the stability criterion. This is true
even if the prompt detonation is suppressed: the delayed detonation also is
numerically unstable. For comparison purposes, we developed an option for our
burner to use a ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied\footnote{To achieve this we directly multiply the
right-hand-side vector in the integration by a factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn.}. When we
evolve the baseline case with this suppressed burner, the prompt detonation
is indeed suppressed. A later detonation does occur, but this delayed
detonation also triggers the suppression. The result is that the detonation
does not get to very high temperatures, and the result is again a ``failed''
supernova that does not generate nearly enough nickel. Thus we do not
reproduce the claim of \citeauthor{kushnir:2013} that the suppressed
burning mode suppresses only the prompt detonation, but allows the
delayed detonation to occur. We could find some way to turn off the
suppression and allow the delayed detonation, but without some independent
confirmation that the delayed detonation is physical (while the prompt
detonation is not) this would be begging the question.

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017} indicates that typical critical
radii for detonation of equal C/O material are in the range 1-10 km.
So, as pointed out by \citeauthor{kushnir:2013}, this numerical
stability issue is mainly a concern for simulations with
resolution of O(10 km) or worse. Detonations observed at higher
resolution are physically plausible because the critical temperature
gradient will be resolved. We have developed an AMR criterion
to help achieve this resolution: we tag on regions suggested by
the burning stability criterion, \autoref{eq:burning_limiter_2}.
The original motivation for this choice was that by lowering
$\Delta x$ and thus the sound-crossing time, we could reach
the point where the detonation was occurring in a numerically stable
region. This goal was not realized due to the reasons discussed above.
Instead, we simply use it as a guide for where to refine. We tag
for refinement all zones with $10^{-6} \leq f_{s} \leq 1$, where
$f_{s}$ is the ratio of the sound crossing time to the energy injection
time. We do not target zones for refinement with $f_{s} > 1$ since
at that point a detonation is locked in. We refine up to a specified
maximum resolution ratio relative to the coarse grid.

>>>>

It is true that when our resolution is low enough, there is a floor on the size
of a hotspot, possibly making such a detonation more likely. However, this is an
unavoidable consequence of the low resolution: it is the correct result of the
simulation that was performed. If the results do not match what occurs at higher
resolution, then the simulation is not converged and the results are not reliable.
However, it may also be the case that a higher resolution simulation will yield
similar results, for example because even at the higher resolution, the physical
size of the hotspot stays the same. Additionally, one might think that simply
adding adaptive mesh refinement in the burning region could help, because then
$\Delta x$ decreases, so the ratio of $\tau_s$ to $\tau_e$ should decrease.
We developed an AMR criterion to do just this: we tag on regions suggested by
the burning stability criterion, \autoref{eq:burning_limiter_2}. But
if a low resolution simulation develops a very large hotspot that violates the
stability criterion, adding refinement inside this region will likely not help,
because the energy trapped in the center of the region may not be able to escape
before ignition occurs. So for low resolution simulations, even adding a significant
amount of refinement (say, refining by a factor of 64 or 128) did not change the
propagation of a detonation.

For the above reasons, an appeal to the numerical instability criterion alone is
insufficient to understand whether a given ignition is real. We find that \textit{all}
of the detonations we have seen at low and moderate resolution inevitably violate
the stability criterion. This is true even if the prompt detonation is suppressed:
the delayed detonation also is numerically unstable. For comparison purposes, we
developed an option for our burner to use a ``suppressed'' burning mode. In a
suppressed burn, we limit the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied\footnote{To achieve this we directly multiply the
right-hand-side vector in the integration by a factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn.}. When we
evolve the baseline case with this suppressed burner, the prompt detonation
is indeed suppressed. A later detonation does occur, but this delayed
detonation also triggers the suppression. The result is that the detonation
does not get to very high temperatures, and the result is again a ``failed''
supernova that does not generate nearly enough nickel. Thus we do not
reproduce the claim of \citeauthor{kushnir:2013} that the suppressed
burning mode suppresses only the prompt detonation, but allows the
delayed detonation to occur. We could find some way to turn off the
suppression and allow the delayed detonation, but without some independent
confirmation that the delayed detonation is physical (while the prompt
detonation is not) this would be begging the question.

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017} indicates that typical critical
radii for detonation of equal C/O material are in the range 1-10 km.
So, as pointed out by \citeauthor{kushnir:2013}, this numerical
stability issue is mainly a concern for simulations with
resolution of O(10 km) or worse. Detonations observed at higher
resolution are physically plausible because the critical temperature
gradient will be resolved. The high resolution needs to be achieved
with some combination of high resolution of the collision process
itself, and perhaps AMR based on the stability criterion as described
above, where we tag for refinement all zones above some $f_{s}$, the
ratio of the sound crossing time to the energy injection time. It
cannot be solely based on the burning stability criterion for the reason
described above, that a detonation may be spuriously locked in by
the low resolution regardless of how much resolution we add to it.
